Surface Source Extension, Reflection Case

### 2019-12-16 repro/papers/sserefl/paper.tex
prelim section

This paper will use linear acoustics in velocity-pressure form as its
model for wave motion. In this model, the space-time fields $p$ is excess
pressure, $\bv$ is particle velocity about equilibrium, $f$
represents an energy source as a defect in the acoustic constitutive
law. The bulk modulus $\kappa$ and density $\rho$ are positive functions of
spatial position. The acoustic wave system couples these quantities:
\begin{eqnarray}
\label{eqn:awe}
\frac{\partial p}{\partial t} + \kappa \nabla \cdot {\bf v} &=&
                                                                f \nonumber\\
\rho \frac{\partial {\bf v}}{\partial t} + \nabla p = 0 \nonumber\\
p,{\bf v}&=&0, t \ll 0
\end{eqnarray}
In terms of the solution of the system \ref{eqn:awe}, the modeling
operator is defined by
\begin{equation}
\label{eqn:fwd}
S[\kappa]f = p|_{z=z_r}
\end{equation}
Two special types of right-hand side are important in the discussion
to follow: isotropic point sources, of the form
\begin{equation}
  \label{eqn:ipr}
  f(\bx,t;\bx_s) = w(t)\delta(\bx-\bx_s).  
\end{equation}
and surface sources, of the form
\begin{equation}
  \label{eqn:ss}
  f(\bx,t;\bx_s) = u(x,y,t;x_s,y_s)\delta(z-z_s). 
\end{equation}
Point sources are a special case of surface sources. Introduce appropriately
parametrized versions of the modeling operator for these two cases:
\begin{eqnarray}
 f(\bx,t;\bx_s) = w(t)\delta(\bx-\bx_s) & \Rightarrow & S[\kappa]f = S_p
  [\kappa]w\\ \nonumber
  \label{eqn:specsrc}
f(\bx,t;\bx_s) = u(x,y,t;x_s,y_s)\delta(z-z_s) & \Rightarrow & S[\kappa]f =
  S_s[\kappa]u
\end{eqnarray}

@@

\subsubsection{Surface Source Extension}
Beside the reformulation of WRI described above, it is also possible
to avoid the storage problem of WRI by limiting the support of the
extended source $f$ to lower dimensional sets. Of several
possibilities discussed in the literature, I will discuss the 
  surface source extension approach introduced by
\cite{HuangNammourSymesDollizal:SEG19}. The extension in question is
defined in equation \ref{eqn:defss}, and denoted $S_s$. The same
annihilator \ref{eqn:defann} is used in this approach, this time
interpreted as acting in the plane $z=z_s$, that is,
\[
  Au(x,y,t;x_s,y_s) = \sqrt{(x-x_s)^2+(y-y_s)^2}u(x,y,t;x_s,y_s).
\]
The problem formulation is
\begin{equation}
  \label{eqn:defsse}
  \mbox{Given } d, \mbox{ find }\kappa \mbox{ and }u \mbox{ to
    minimize }
  \| d -  S_s[\kappa]u\|^2+\lambda^2\|Au\|^2.
\end{equation}
Evidently the extended source has as many parameters as the data, so
the memory complexity of this problem is comparable to that of FWI or
reformulated WRI. The computational complexity compares favorably to
that of reformulated WRI.

\subsection{Layered Media, Plane Waves}
As explained in the introduction, this study explores the properties
of the various FWI variants explained above for the special case of
layered media: $\kappa = \kappa(z)$,
$\rho=\rho(z)$. Then the modeled shots differ from
each other by a horizontal translation, so that there is in effect
only one shot, for which I take the location to be
$\bx_s=(0,0,0)$ and suppress $\bx_s$ from the notation.

If $\log \kappa$ and $\log \rho$ are globally bounded and $f$ has
compact support (and some other properties that will be made precise
later), there is $\xi_{\rm max} > 0$ so that for
$\|\xi\|<\xi_{\rm max}$, the integrals defining slant stack fields
$\tilde{p},\tilde{\bv}$, namely
\begin{eqnarray}
\label{eqn:slant}
\tp(\xi,z,t)& = &\int dx\, dy\, p(x,y,z,t+\xi x),\nonumber\\
\tv(\xi,z,t)&=&\int dx\, dy\, v_z(x,y,z,t+\xi x),\nonumber\\
\tf(\xi,z,t)&=&\int dx\,dy\, f(x,y,z,t+\xi x),
\end{eqnarray}
are absolutely convergent, and $\tp, \tv$ satisfy the plane wave
version of the acoustic system \ref{eqn:awe}:
\begin{eqnarray}
\label{eqn:awepw}
(1-c^2\xi^2)\frac{\partial \tp}{\partial t} + \kappa\frac{\partial
  \tv}{\partial z} &=& \tf \nonumber\\
\rho \frac{\partial \tv}{\partial t} + \frac{\partial \tp}{\partial
  z}&=&0\nonumber\\
\tp,\tv&=&0, t \ll 0.
\end{eqnarray}
Here $c=\sqrt{\kappa/\rho}$ is the sound velocity. I use $\xi$ instead
of the more usual $p$ for plane wave slowness since $p$ already has
been assigned to mean pressure. Note that for a isotropic point
source, respectively a surface source,
\begin{eqnarray}
  f(\bx,t;\bx_s)=w(t)\delta(\bx) & \Rightarrow &
                                               \tf(\xi,z,t)=w(t)\delta(z)\\ \nonumber 
  f(\bx,t;\bx_s) = u(x,y,t)\delta(z) & \Rightarrow &
                                                     \tf(\xi,z,t) =
                                                     \tu(\xi,t)\delta(z)\\
  \nonumber
  \tu(\xi,t) &=& \int dx\,dy\, u(x,y,t+\xi x)
\end{eqnarray}
The plane wave domain modeling
operator for space-time extended sources samples $\tp$ on $z=z_r$:
\begin{equation}
\label{eqn:modpw}
\tS[\kappa]\tf = \tp|_{z=z_r}
\end{equation}
Similarly, the plane wave domain versions of $S_p$ and $S_s$ are
\begin{eqnarray}
 \tf(\xi,z,t) = w(t)\delta(z) & \Rightarrow & \tS[\kappa]\tf = \tS_p
  [\kappa]w\\ \nonumber
  \label{eqn:specsrc}
\tf(\xi,z,t) = \tu(\xi,t)\delta(z-z_s) & \Rightarrow & \tS[\kappa]\tf =
  \tS_s[\kappa]\tu
\end{eqnarray}

For homogeneous $\kappa=\kappa_0, \rho=\rho_0, c=c_0$, a computation
recapitulated in the Appendix C, relying on computations for the 1D
acoustics system detailed in Appendices A and B, yields
\begin{equation}
\label{eqn:modhom}
\tS_s[\kappa_0]\tu(\xi,t) = \frac{1}{2(1-c_0^2\xi^2)}\tu
\left(\xi,t-z_r\sqrt{\frac{1}{c_0^2}-\xi^2}\right).
\end{equation}
That is, at a homogeneous model, the plane wave modeling operator is
simply scaling by a $\xi$-dependent factor.

\section{Main Results}

@@

Assume that $f$ is square-integrable, $\kappa = \kappa(z)$,
$\rho=\rho(z)$ (layered medium). Then the modeled shots differ from
each other by a horizontal translation, so that there is in effect
only one shot, for which I take the location to be
$\bx_s=(0,0,z_s)$ and suppress $\bx_s$ from the notation.

@@

 If $\log \kappa$ and $\log \rho$ are globally
bounded, there is $\xi_{\rm max} > 0$ so that for $\|\xi\|<\xi_{\rm
  max}$, the integrals defining slant stack fields
$\tilde{p},\tilde{\bv}$, namely
\begin{eqnarray}
\label{eqn:slant}
\tp(\xi,z,t)& = &\int dx p(x,z,t+\xi x),\nonumber\\
\tv(\xi,z,t)&=&\int dx \bv(x,z,t+\xi x),\nonumber\\
\tf(\xi,z,t)&=&\int dx f(x,z,t+\xi x),
\end{eqnarray}
are absolutely convergent, and $\tp, \tv$ satisfy 
\begin{eqnarray}
\label{eqn:awepw}
(1-c^2\xi^2)\frac{\partial \tp}{\partial t} + \kappa\frac{\partial
  \tv}{\partial z} &=& \tf \nonumber\\
\rho \frac{\partial \tv}{\partial t} + \frac{\partial \tp}{\partial
  z}&=&0\nonumber\\
\tp,\tv&=&0, t \ll 0.
\end{eqnarray}
Here $c=\sqrt{\kappa/\rho}$ is the sound velocity. I use $\xi$ instead
of the more usual $p$ for plane wave slowness since $p$ already has
been assigned to mean pressure. Note that if
$f(\bx,t;\bx_s)=w(t)\delta(\bx)$, then
$\tf(\xi,z,t)=w(t)\delta(z)$. Likewise for a surface extended source,
$f(\bx,t;\bx_s) = w(x,t)\delta(z)$, $\tf(\xi,z,t) = \tw(\xi,t)\delta(z)$.

From now on assume a surface extended source, so in plane wave domain
$\tf(\xi,z,t) = \tw(\xi,t)\delta(z)$. The plane wave domain modeling
operator samples $\tp$ on $z=z_r=0$:
\begin{equation}
\label{eqn:modpw}
\tS[\kappa]\tw = \tp|_{z=0}
\end{equation}

@@

\append{1D Radiation Problem}
Begin with the 1D acoustics point source system. 
\begin{eqnarray}
\label{eqn:awe1d}
\frac{\partial p}{\partial t} +\kappa\frac{\partial 
  v}{\partial z} &=& w(t)\delta(z-z_s) \nonumber\\
\rho \frac{\partial v}{\partial t} + \frac{\partial p}{\partial 
  z}&=&0\nonumber\\
 p,v&=&0, t \ll 0. 
\end{eqnarray}
Since the right hand side is singular, so is the solution, so it must
be a solution in the weak sense; since it is causal, it is uniquely
determined by the system above. It follows from the weak solution
conditions that the pressure is continuous at $p=0$, whence $v$ must
have a discontinuity.

>>>>

This paper will use linear acoustics in velocity-pressure form as its
model for wave motion. In this model, the space-time fields $p$ is excess
pressure, $\bv$ is particle velocity about equilibrium, $f$
represents an energy source as a defect in the acoustic constitutive
law. The bulk modulus $\kappa$ and density $\rho$ are positive functions of
spatial position. The acoustic wave system couples these quantities:
\begin{eqnarray}
\label{eqn:awe}
\frac{\partial p}{\partial t} + \kappa \nabla \cdot {\bf v} &=&
                                                                f \nonumber\\
\rho \frac{\partial {\bf v}}{\partial t} + \nabla p = 0 \nonumber\\
p,{\bf v}&=&0, t \ll 0
\end{eqnarray}
In terms of the solution of the system \ref{eqn:awe}, the modeling
operator is defined by
\begin{equation}
\label{eqn:fwd}
S[\kappa]f = p|_{z=z_r}
\end{equation}
Two special types of right-hand side are important in the discussion
to follow: isotropic point sources, of the form
\begin{equation}
  \label{eqn:ipr}
  f(\bx,t;\bx_s) = w(t)\delta(\bx-\bx_s).  
\end{equation}
and surface sources, of the form
\begin{equation}
  \label{eqn:ss}
  f(\bx,t;\bx_s) = u(x,y,t;x_s,y_s)\delta(z-z_s). 
\end{equation}
Point sources are a special case of surface sources. Introduce appropriately
parametrized versions of the modeling operator for these two cases:
\begin{eqnarray}
  \label{eqn:defpt}
 f(\bx,t;\bx_s) = w(t)\delta(\bx-\bx_s) & \Rightarrow & S[\kappa]f = S_p
  [\kappa]w\\ 
  \label{eqn:defss}
f(\bx,t;\bx_s) = u(x,y,t;x_s,y_s)\delta(z-z_s) & \Rightarrow & S[\kappa]f =
  S_s[\kappa]u
\end{eqnarray}

@@

\subsubsection{Surface Source Extension}
Beside the reformulation of WRI described above, it is also possible
to avoid the storage problem of WRI by limiting the support of the
extended source $f$ to lower dimensional sets. Of several
possibilities discussed in the literature, I will discuss the 
  surface source extension approach introduced by
\cite{HuangNammourSymesDollizal:SEG19}. The extension in question is
defined in equation \ref{eqn:defss}, and denoted $S_s$. The same
annihilator \ref{eqn:defann} is used in this approach, this time
interpreted as acting in the plane $z=z_s$, that is,
\[
  Au(x,y,t;x_s,y_s) = \sqrt{(x-x_s)^2+(y-y_s)^2}u(x,y,t;x_s,y_s).
\]
or
\[
  Au(x,t;x_s) = (x-x_s)u(x,t;x_s)
\]
for 2D. 
The problem formulation is
\begin{equation}
  \label{eqn:defsse}
  \mbox{Given } d, \mbox{ find }\kappa \mbox{ and }u \mbox{ to
    minimize }
  \| d -  S_s[\kappa]u\|^2+\lambda^2\|Au\|^2.
\end{equation}
Evidently the extended source has as many parameters as the data, so
the memory complexity of this problem is comparable to that of FWI or
reformulated WRI. The computational complexity compares favorably to
that of reformulated WRI.

\subsection{Layered Media, Plane Waves}
As explained in the introduction, this study explores the properties
of the various FWI variants explained above for the special case of
layered media: $\kappa = \kappa(z)$,
$\rho=\rho(z)$. Then the modeled shots differ from
each other by a horizontal translation, so that there is in effect
only one shot, for which I take the location to be
$\bx_s=(0,0,z_s)$.

If $\log \kappa$ and $\log \rho$ are globally bounded and $f$ has
compact support (and some other properties that will be made precise
later), there is $\xi_{\rm max} > 0$ so that for
$\|\xi\|<\xi_{\rm max}$, the integrals defining slant stack fields
$\tilde{p},\tilde{\bv}$, namely
\begin{eqnarray}
\label{eqn:slant}
\tp(\xi,z,t)& = &\int dx\, dy\, p(x,y,z,t+\xi x),\nonumber\\
\tv(\xi,z,t)&=&\int dx\, dy\, v_z(x,y,z,t+\xi x),\nonumber\\
\tf(\xi,z,t)&=&\int dx\,dy\, f(x,y,z,t+\xi x),
\end{eqnarray}
are absolutely convergent, and $\tp, \tv$ satisfy the plane wave
version of the acoustic system \ref{eqn:awe}:
\begin{eqnarray}
\label{eqn:awepw}
(1-c^2\xi^2)\frac{\partial \tp}{\partial t} + \kappa\frac{\partial
  \tv}{\partial z} &=& \tf \nonumber\\
\rho \frac{\partial \tv}{\partial t} + \frac{\partial \tp}{\partial
  z}&=&0\nonumber\\
\tp,\tv&=&0, t \ll 0.
\end{eqnarray}
Here $c=\sqrt{\kappa/\rho}$ is the sound velocity. I use $\xi$ instead
of the more usual $p$ for plane wave slowness since $p$ already has
been assigned to mean pressure. Note that for a isotropic point
source, respectively a surface source,
\begin{eqnarray}
  f(\bx,t;\bx_s)=w(t)\delta(\bx) & \Rightarrow &
                                               \tf(\xi,z,t)=w(t)\delta(z)\\ \nonumber 
  f(\bx,t;\bx_s) = u(x,y,t)\delta(z) & \Rightarrow &
                                                     \tf(\xi,z,t) =
                                                     \tu(\xi,t)\delta(z)\\
  \nonumber
  \tu(\xi,t) &=& \int dx\,dy\, u(x,y,t+\xi x)
\end{eqnarray}
The plane wave domain modeling
operator for space-time extended sources samples $\tp$ on $z=z_r$:
\begin{equation}
\label{eqn:modpw}
\tS[\kappa]\tf = \tp|_{z=z_r}
\end{equation}
Similarly, the plane wave domain versions of $S_p$ and $S_s$ are
\begin{eqnarray}
  \label{eqn:pwpt}
 \tf(\xi,z,t) = w(t)\delta(z-z_s) & \Rightarrow & \tS[\kappa]\tf = \tS_p
  [\kappa]w\\ 
  \label{eqn:pwss}
\tf(\xi,z,t) = \tu(\xi,t)\delta(z-z_s) & \Rightarrow & \tS[\kappa]\tf =
  \tS_s[\kappa]\tu
\end{eqnarray}

For homogeneous $\kappa=\kappa_0, \rho=\rho_0, c=c_0$, a computation
recapitulated in the Appendix C, relying on computations for the 1D
acoustics system detailed in Appendices A and B, yields
\begin{eqnarray}
  \label{eqn:ptmodhompw}
  \tS_p[\kappa_0]w(\xi,t) & = &\frac{1}{2(1-c_0^2\xi^2)}w
                                \left(t-\frac{|z_r-z_s|}{\tc_0(\xi)}\right)\\
    \label{eqn:ssmodhompw}
  \tS_s[\kappa_0]\tu(\xi,t) & = &\frac{1}{2(1-c_0^2\xi^2)}\tu 
                                  \left(\xi,t-\frac{|z_r-z_s|}{\tc_0(\xi)}\right)\\
    \label{eqn:stmodhompw}
  \tS[\kappa_0]\tf(\xi,t) & = &  \frac{1}{2}(1-c_0^2\xi^2)^{-1}\int dz \tf\left(\xi,z,t -
                                \frac{|z_r-z|}{\tc_0(\xi)}\right)                              
\end{eqnarray}
Here
\begin{equation}
  \label{eqn:vertvel}
  \tc(\xi,z) = \frac{c(z)}{\sqrt{1-c(z)^2\xi^2}}
    \end{equation}
is the vertical plane wave velocity, and we write
$\tc_0(\xi)=\tc(\xi,z)$ for any $z$, when $c=c_0$.

\subsection{Extended FWI in Plane Wave Domain}
Of the various extended FWI formulation explained above, three do not
require the storage of full space-time fields:
\begin{itemize}
\item Least-squares FWI;
\item Reformulated WRI; and
\item Surface Source Extension.
\end{itemize}
The remainder of this paper is devoted to analysis of these three
approaches.

It is possible to transform the objectives posed in the space-time
domain problem statements above, but this leads to technical
complications that do not affect the conclusions. Therefore I pose
analogous problems directly in the plane wave domain. The $L^2$ norms
(integral root mean squares) below involve integration over a range of
plane wave slownesses $\xi$, which for simplicity I assume symmetric,
that is, $-\xi_{\rm max} \le \xi \le \xi_{\rm max}$, and over
appropriate ranges of other variables which I do not state explicitly
here (see the analysis sections later in the paper where all such
choices are made explicitly). As before, the $L^2$-norm squared symbol
$\|\cdot\|^2$ means the integral over all variables of the square of
whatever is inside.

\subsubsection{Least Squares FWI}

\begin{equation}
  \label{eqn:defolspw}
  \mbox{Given } \td, \mbox{ find }\kappa \mbox{ and }w \mbox{ to
    minimize }
  \| \td -  \tS_p[\kappa]w\delta(\cdot-z_s)\|^2.
\end{equation}

\subsubsection{Reformulated WRI}

\begin{equation}
  \label{eqn:defwrireformpw}
  \mbox{Given } d, \mbox{ find }\kappa \mbox{ and }e \mbox{ to
    minimize }
  \| \td -  \tS[\kappa]w_*\delta(\cdot-z_s) - \tS[\kappa]\tS[\kappa]^Te\|^2+\lambda^2\|\tS[\kappa]^Te\|^2.
\end{equation}

The reformulated WRI objective involves 
the product $\tS[\kappa]\tS[\kappa]^T$ in which $\tS[\kappa]^T$ is the
formal or $L^2$ adjoint. For the homogeneous case,
$\kappa=\kappa_0,\rho=\rho_0,c=c_0$,
\begin{equation}
  \label{eqn:stmodhompwtrans}
  (\tS[\kappa]^T\te)(\xi,z,t)=\frac{1}{2}(1-c_0^2\xi^2)^{-1}\te\left(\xi,t +
    \frac{|z_r-z|}{\tc_0(\xi)}\right).
\end{equation}

While I have endeavoured to limit mathematical technicalities in this background
presentation, one such technicality intrudes at this point: the
output of the transpose operator $\tS[\kappa]^T$ is nonzero and non-decaying arbitrarily
far in space and in negative time. As will be obvious shortly, this
fact prevents the composition $\tS[\kappa]\tS[\kappa]^T$ from being
well-defined, without some further constraint being imposed on the
domain of $\tS[\kappa]$. The simplest appears to be the assumption
that all plane wave sources act over a limited depth range, $0 \le z
\le z_{\rm max}$. Then the $z$ integrals appearing in the expressions
for $S$ and $\tS$ have limits $[0,z_{\rm max}]$

With this amendment, compute the composition:
\[
  \tS[\kappa]\tS[\kappa]^T\te(\xi,t) = \frac{1}{2}(1-c_0^2\xi^2)^{-1}
\]
\[
  \times 
  \int_0^{z_{\rm max}} dz\, \frac{1}{2}(1-c_0^2\xi^2)^{-1}\te\left(\xi,t' +
    \frac{|z_r-z|}{\tc_0(\xi)}\right)_{t'=t-\frac{|z_r-z|}{\tc_0(\xi)}}
\]
\begin{equation}
  \label{eqn:stmodhompwnorm}
 = \frac{z_{\rm max}}{4}(1-c_0^2\xi^2)^{-2}\te(\xi,t)
\end{equation}

\subsubsection{Surface Source Extension}

The surface source space-time formulation employs an
annihilator $A$, defined in \ref{eqn:defste}. The corresponding
operator in the plane wave domain is somewhat easier to compute for the 2D
case: one obtains 
\[
\tA = D_{\xi}I_t,
\]
that is, differentiation in $\xi$ and integration in $t$, in either
order. The 3D expression is somewhat more complicated. Since the
examples below will be computed in 2D, the above definition suffices.

The surface source extended FWI problem in the plane wave domain is then
\begin{equation}
  \label{eqn:defssepw}
  \mbox{Given } \td, \mbox{ find }\kappa \mbox{ and }\tu \mbox{ to
    minimize }
  \| \td -  \tS_s[\kappa]\tu\|^2+\lambda^2\|\tA\tu\|^2.
\end{equation}

\subsection{Inversion Velocity Analysis: the zero penalty limit}
Examination of the operators appearing in the extended inversion
problems considered here (equations \ref{eqn:stmodhompwnorm} for
reformulated WRI, \ref{eqn:ssmodhompw} for SSE) reveals that they
are both invertible. This property permits a simplification of the
penalty formulations \ref{eqn:defwrireformpw} and
\ref{eqn:defssepw}. This simplification has been used several times in
prior work on extended FWI
\cite[]{Symes:EAGE15,Herve2017,HouSymes:Geo18}, and has come to be
called Inversion Velocity Analysis (IVA) in the exploration seismology
literature.

The idea is well-known in numerical optimization, and can most easily
be explained in the context of the SSE objective function, in
plane-wave domain:
\begin{equation}
  \label{eqn:ssepen}
  \kappa,\tu \mapsto \|\td-\tS_s[\kappa]\tu\|^2 + \lambda^2\|\tA\tu\|^2
\end{equation}

The transformation to IVA takes place in two steps. The first is
trivial: simply divide the objective function by $\lambda^2$, to
obtain
\begin{equation}
  \label{eqn:ssepeninv}
  \kappa,\tu \mapsto \lambda^{-2}\|\td-\tS_s[\kappa]\tu\|^2 + \|\tA\tu\|^2
\end{equation}
Only the minimum values change: for any particular $\lambda \in
\bR^+$, the sets of local and global minimizers are exactly the same.

The second step in moving from this objective to the corresponding IVA
objective is of value in itself: it consists of eliminating the linear
variable $\tu$ by minimizing the function defined in \ref{eqn:ssepen}
over $\tu$. The resulting minimum is a function of $\kappa$: {\em a
  priori}, a local minimizer is the same as the $\kappa$ component of
a local minimizer of the original objective. This reduction of the
objective to a function of some of its variables is known as the {\em
  Variable Projection Method} (VPM), especially if the eliminated variables
appear linearly in the original problem, as they do here
\cite[]{GolubPereyra:03}. For the plane wave surface source objective,
the result of VPM reduction is
\begin{equation}
  \label{eqn:ssevpm}
  \kappa \mapsto \lambda^{-2}\|\td-\tS_s[\kappa]\tu[\kappa]\|^2 + \|\tA\tu[\kappa]\|^2
\end{equation}
in which $\tu[\kappa]$ is the minimizer of the base objective
\ref{eqn:ssepeninv} over $\tu$, with $\kappa$ held fixed:
$\tu[\kappa]$ solves the {\em normal equation}
\begin{equation}
  \label{eqn:ssenorm}
  (\lambda^{-2}\tS_s[\kappa]^T\tS_s[\kappa] +
  \tA^T\tA)\tu[\kappa]=\lambda^{-2}\tS_s[\kappa]^T\td.
\end{equation}

The major import of VPM is that the gradient
of the reduced objective, a key element in any calculus-based
optimization algorithm, is relatively simple to compute: formally, it
is simply the gradient of the original objective with the
to-be-eliminated variables held fixed. See \cite{GolubPereyra:03} for
an explanation, and
\cite{vanLeeuwenMulder:09,Rickett:SEG12,LiRickettAbubakar:13} for
applications to geophysical inversion.

The final step is to observe that the VPM objective converges as
$\lambda \rightarrow 0$. Examination of \ref{eqn:ssepeninv} suggests
that for this to happen, the data must be fit perfectly, at least in
the limit. This is likely to happen only if the linear modeling
operator ($\tS_s[\kappa]$, in this case) is invertible - which it is,
at least in the case of constant $\kappa=\kappa_0$,
see equation \ref{eqn:ssmodhompw}.  In fact $\tS[\kappa]$ is a
continuous function of $\kappa$, as is explained in Appendix \appSingular, so in fact data is fit exactly and by precisely one $\tu$ for
all $\kappa$ sufficiently close to constant. It is not hard to see that the
$\lambda = 0$ limit of the VPM objective obtained from 
\ref{eqn:ssepeninv} is
\begin{equation}
  \label{eqn:ssiva}
  \kappa \mapsto \|A \tS_s[\kappa]^{-1}\td\|^2
\end{equation}
That is, the IVA objective is the mean square of the annihilator ouput
from the necessarily unique extended source field $\tu =
\tS_s[\kappa]^{-1}\td$ that precisely fits the data.

Appendix \appIVA  presents a proof that equation \ref{eqn:ssiva}
actually describes the $\lambda \rightarrow 0$ limit. The version of
surface source extended FWI whose properties are described in the next
section is the minimization problem for this objective, to which I
give a name:
\begin{equation}
  \label{eqn:defssiva}
  J_{\rm SSE} = \frac{1}{2}\|A\tu[\kappa]\|^2, \,
  \tu[\kappa]=\tS_s[\kappa]^{-1}\td
\end{equation}

The story for reformulated WRI is similar. From equation
\ref{eqn:defwrireformpw}, the VPM objective after scaling by
$\lambda^{-2}$ is 
\begin{equation}
  \label{eqn:wrivpm}
  \kappa, \te \mapsto \lambda^{-2}\| \td -  \tS[\kappa]w_*\delta(\cdot-z_s) -
  \tS[\kappa]\tS[\kappa]^T\te[\kappa]\|^2+\|\tS[\kappa]^T\te[\kappa]\|^2,
\end{equation}
in which $\te[\kappa]$ solves the normal equation
\begin{equation}
  \label{eqn:wrinorm}
  (\lambda^{-2}(\tS[\kappa]\tS[\kappa]^T)^2 +
  \tS[\kappa]\tS[\kappa]^T)\te[\kappa] =
  \lambda^{-2}\tS[\kappa]\tS[\kappa]^T(\td -
  \tS[\kappa]w_*\delta(\cdot-z_s) )
\end{equation}
If $\kappa$ is close to $\kappa_0$, then $\tS[\kappa]\tS[\kappa]^T$ is
invertible, as follows from equation \ref{eqn:stmodhompwnorm} and the
continuity of $\tS[\kappa]$ with respect to $\kappa$, mentioned
above. Therefore the normal equation is equivalent to
\begin{equation}
  \label{eqn:wrinorm1}
  (\lambda^{-2}\tS[\kappa]\tS[\kappa]^T +
  I)\te[\kappa] =
  \lambda^{-2}(\td -
  \tS[\kappa]w_*\delta(\cdot-z_s) )
\end{equation}
The $\lambda \rightarrow 0$ limit of the resulting function of
$\kappa$ is
\[
  J_{\rm WRI}[\kappa] = \frac{1}{2}\|\tS[\kappa]^T((\tS[\kappa]\tS[\kappa]^T)^{-1}(\td -
  \tS[\kappa]w_*\delta(\cdot-z_s)))\|^2
\]
\begin{equation}
  \label{eqn:wripwiva}
  = \frac{1}{2}\langle \td -
  \tS[\kappa]w_*\delta(\cdot-z_s), (\tS[\kappa]\tS[\kappa]^T)^{-1}( \td -
  \tS[\kappa]w_*\delta(\cdot-z_s)) \rangle
\end{equation}

Finally, it is entirely possible to apply VPM to the least squares FWI
objective \ref{eqn:defolspw}, as has been suggested as an algorithmic
approach to determining source parameters as part of FWI
\cite{Rickett:SEG12,LiRickettAbubakar:13}. However in this instance it
is more relevant to pose the known-source version of least squares
FWI, as most parallel to WRI (in any of its forms): recalling the
identity \ref{eqn:pwpt}, 
\[
  J_{\rm FWI}[\kappa] = \frac{1}{2}\|\td-\tS_p[\kappa]w_*\|^2
\]
\begin{equation}
  \label{eqn:olspwobj}
  = \frac{1}{2}\langle \td -
  \tS[\kappa]w_*\delta(\cdot-z_s), \td -
  \tS[\kappa]w_*\delta(\cdot-z_s)\rangle.
\end{equation}
Comparing equations \ref{eqn:wripwiva} and \ref{eqn:olspwobj}, reach
the surprising conclusion that
\begin{quote}
  The $\lambda \rightarrow 0$ limit of the WRI-VPM objective
  \ref{eqn:wrivpm} is a weighted-norm variant of the FWI objective
  \ref{eqn:olspwobj}.
\end{quote}
This conclusion is surprising because one would normally expect the
opposite limit, that is $\lambda \rightarrow \infty$, of the WRI-VPM
objective \ref{eqn:wrivpm} to be equivalent in some sense to the FWI
objective \ref{eqn:olspwobj}, rather than the $\lambda \rightarrow 0$
limit. Indeed, the limit $\lambda \rightarrow \infty$ can only make
sense if $\tS[\kappa]\tS[\kappa]^T\te \rightarrow 0$: since
$\tS[\kappa]\tS[\kappa]^T$ is invertible for $\kappa$ near $\kappa_0$,
in that case $\te \rightarrow 0$ necessarily, so the objective
\ref{eqn:wrivpm} presumably tends to \ref{eqn:olspwobj}. This conclusion
is in fact correct, as follows from a variant of the argument
presented in Appendix \appIVA.

The import of the reasoning presented in Appendix \appIVA is that for
$\lambda$ sufficiently small, the behaviour of the VPM objective in
each of these cases is close to that of the $\lambda \rightarrow 0$
limit. As is well-known, and will be recapitulated in the Main Results
section, the least squares FWI objective does not generate
constructive model updates. Since all of the VPM objectives explored
here become close to the least squares FWI objective for large
$\lambda$, it is the case of small $\lambda$ that holds any possible
promise of generating substantially more constructive updates, and we
will see below that indeed in some cases that promise is realized. The
$\lambda \rightarrow 0$ limit provides a convenient window into
small $\lambda$ behaviour, so the focus in the remainder of the paper
will be on the limiting problems defined in \ref{eqn:olspwobj},
\ref{eqn:wripwiva}, and \ref{eqn:defssiva}.
\section{Main Results}

@@

Assume that $\kappa = \kappa(z)$,
$\rho=\rho(z)$ (layered medium). Then the modeled shots differ from
each other by a horizontal translation, so that there is in effect
only one shot, for which I take the location to be
$\bx_s=(0,0,z_s)$ and suppress $\bx_s$ from the notation.

@@

 If $\log \kappa$ and $\log \rho$ are globally
bounded, and $f$ is square-integrable, there is $\xi_{\rm max} > 0$ so that for $\|\xi\|<\xi_{\rm
  max}$, the integrals defining slant stack fields
$\tilde{p},\tilde{\bv}$, namely
\begin{eqnarray}
\label{eqn:slant}
\tp(\xi,z,t)& = &\int dx p(x,z,t+\xi x),\nonumber\\
\tv(\xi,z,t)&=&\int dx \bv(x,z,t+\xi x),\nonumber\\
\tf(\xi,z,t)&=&\int dx f(x,z,t+\xi x),
\end{eqnarray}
are absolutely convergent, and $\tp, \tv$ satisfy 
\begin{eqnarray}
\label{eqn:awepw}
(1-c^2\xi^2)\frac{\partial \tp}{\partial t} + \kappa\frac{\partial
  \tv}{\partial z} &=& \tf \nonumber\\
\rho \frac{\partial \tv}{\partial t} + \frac{\partial \tp}{\partial
  z}&=&0\nonumber\\
\tp,\tv&=&0, t \ll 0.
\end{eqnarray}
Here $c=\sqrt{\kappa/\rho}$ is the sound velocity. I use $\xi$ instead
of the more usual $p$ for plane wave slowness since $p$ already has
been assigned to mean pressure. Note that if
$f(\bx,t;\bx_s)=w(t)\delta(\bx)$, then
$\tf(\xi,z,t)=w(t)\delta(z)$. Likewise for a surface extended source,
$f(\bx,t;\bx_s) = w(x,t)\delta(z)$, $\tf(\xi,z,t) = \tw(\xi,t)\delta(z)$.

From now on assume a surface extended source, so in plane wave domain
$\tf(\xi,z,t) = \tw(\xi,t)\delta(z)$. The plane wave domain modeling
operator samples $\tp$ on $z=z_r$:
\begin{equation}
\label{eqn:modpw}
\tS[\kappa]\tw = \tp|_{z=z_r}
\end{equation}

@@

\append{1D Radiation Problem}
Begin with the 1D acoustics point source system. 
\begin{eqnarray}
\label{eqn:awe1dptsrc}
\frac{\partial p}{\partial t} +\kappa\frac{\partial 
  v}{\partial z} &=& w(t)\delta(z-z_s) \nonumber\\
\rho \frac{\partial v}{\partial t} + \frac{\partial p}{\partial 
  z}&=&0\nonumber\\
 p,v&=&0, t \ll 0. 
\end{eqnarray}
Since the right hand side is singular, so is the solution, so it must
be a solution in the weak sense. It follows from the weak solution
conditions that the pressure is continuous at $p=0$, whence $v$ must
have a discontinuity.
